Generating Empirically Optimized Numerical Software from MATLAB Prototypes

### 2008-09-20 doc/papers/cgo08/paper.tex
submitted version
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@358 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

% CGO 09: http://www.cgo.org/
% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
% Max 22 pages, double-spaced, 11pt or larger font
% Abstract 300 words: Sep. 12, 2008
% Full paper: Sep. 19, 2008

@@

\setlength{\textwidth}{6.45in}
\setlength{\oddsidemargin}{0.05in}
\setlength{\evensidemargin}{0.05in}

\setlength{\topmargin}{-0.5in}
\setlength{\textheight}{9in}

\usepackage{url}
\usepackage{html}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{cite}
\usepackage{amsmath}
%\usepackage[margin=1in]{geometry}

@@

\maketitle
\doublespacing

\abstract{

The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations. Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.

In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.  To generate highly optimized code from a high-level MATLAB prototype, we will follow a three-step approach.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.   We then use the source transformation tool PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously. The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism. Finally, we employ the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment. Orio performs an automated empirical search for selecting the best among the multiple optimized code variants.

We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
}

\section{Introduction}
\label{sec:intro}

@@

Historically, the research community has pursued two separate paths toward the goal of making software run at near-peak levels.  The first of these builds on research into compilers and their associated technologies.  The holy grail of compilation research is to take an arbitrary code as input and produce optimal code as output for a given language and hardware platform.  Despite the immense effort that has been poured into that approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.    The second path has been to identify kernel routines that constitute the dominant costs of a wide variety of applications.  When such kernels can be identified and agreed upon by the members of the community, programmers with the required level of technical knowledge can concentrate on producing these optimized kernel libraries for architectures of interest. A prime example is the high-performance BLAS libraries produced by a combination of hardware vendors, independent software vendors, and researchers.  Developers who write their codes calling these routines can achieve high performance across all supported architectures. The efforts in high-performance library development aptly demonstrate, however, that the research into compiler technologies has not been fully successful.  If compilers could take the relatively simple code even for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.    Like compiler optimization, the kernel library-oriented approach has significant limitations.  Support for a variety of architectures is the most serious problem because the optimizations necessary to achieve near-peak performance are by nature not portable.   Lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.  It is a sad fact of the computing industry that by the time highly-optimized code is available for a given architecture, that architecture is generally well on its way to obsolescence.

We are working on a combination of the two approaches designed to overcome some of their shortcomings.   This paper described our three-pronged approach toward the development of an automatically tuned library for matrix algebra computations.  In Section~\ref{sec:matlab}, we describe the compiler we have developed that converts a MATLAB script into simple C code. In Section~\ref{sec:orio}, we demonstrate how to employ the empirical performance tuning system Orio in conjunction with the PLuTo tool to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.  The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism. In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.

@@

The most common approach  to tuning numerical codes is for an expert to transform the source manually, unrolling loops, blocking for multiple levels of cache, and inserting prefetch instructions.  The pitfalls of this approach are well-understood~\cite{Goedecker01}.  It requires a significant amount of time and effort. Optimizing code for one particular platform may in fact make it less efficient on other platforms and often makes it complex and hard to understand or maintain.    An alternative is the use of tuned libraries of key numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific computing in C++, or GPULib by Tech-X, an open source library for accelerating numerical computations on Graphics Processing Units~\cite{gpulib}, to name just a few). Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:composed}, a series of calls to library routines can incur high memory costs.

@@

The best performance today is obtained through approaches like that taken by PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists. Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated. We propose to investigate ways to capture this performance tuning expertise in a way that can be integrated into the application. We will thus enable performance portability and ensure that the improvements to the application are long-lasting.

A number of source or binary transformation tools for performance-improving optimizations exist. LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.  A new tool, POET~\cite{POET} also supports a number of loop transformations. POET offers a complex template-based syntax for defining transformations in a language-independent manner. Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops. PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron. Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations. The transformations target both improved cache locality and parallelism.

\section{Composed Operations}
\label{sec:composed}

Codes based on matrix algebra are generally constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency. Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g ~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.

The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.  Single routines that carry out more than one linear algebra operation are known as {\t composed} BLAS. As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.   These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies. In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only once. Our preliminary results indicate that this loop fusion leads to a routine that delivers 90\% more megaflops than does a pair of calls to the highly optimized Goto BLAS~\cite{Goto:2006fk} {\tt GEMV} routines for large matrix orders.

Composed routines are the focus of the work presented here, but much of what we discuss generalizes to a much broader array of computations.   Extending our approach will be the subject of future work.

@@

\section{Our Approach}
\label{sec:approach}

To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.  We then (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} developed at Ohio State University to optimize for coarse-grained parallelism and locality simultaneously. Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters. Orio then performs an empirical search to select the best among the multiple optimized code variants.

%Probably need some words about WHY this 3 pronged approach is a good idea.

@@

\caption{Overview of the code generation and tuning process.}
\label{fig:process}
\end{figure}

@@

Orio~\cite{Norris:2007} is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions of the annotated code. Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.

% Explain how Pluto is used

@@

\begin{figure}[thp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.45\textwidth}
\scriptsize
\begin{verbatim}
void vadd(int n, double *y, double *x1,
               double *x2, double *x3)
{
    int i;

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.45\textwidth}
\scriptsize
\begin{verbatim}
spec align_unroll {
 def build  
 { 
   arg build_command = 'mpixlc -O3 -qstrict -lm'; 
 } 
 def performance_counter  
 { 
   arg method = 'basic timer'; 
   arg repetitions = 10000;
 } 
 def performance_params 
 { 
   param UF[] = range(1,20);
 } 
 def input_params 
 { 
   param SIZE = @THESIZE@;
   decl int n = SIZE;
   decl double x1[n] = random;   
   decl double x2[n] = random;   
   decl double x3[n] = random;   
   decl double y[n] = 0;
 } 
 def search
 {
   arg algorithm = 'Exhaustive';
   arg time_limit = 20;
   arg run_command = 'cqsub -n 64 -t 10 -q short ';
   arg num_processes = 64;
 }
}
\end{verbatim}
\end{minipage}\\
\end{tabular}
\caption{Orio example: annotated C source code (left) and a tuning specification for the Blue Gene/P (right).}
\label{fig:orio-example}
\end{figure}

We use the VADD operation, $x = w + y + z$, as a simple example of the annotations and tuning specification used by Orio. Figure~\ref{fig:orio-example} shows the C code with annotation comments indicating that Orio should perform memory alignment and loop transformations. In addition to simple source transformation annotations in the example above, Orio supports a number of other transformations, including loop unroll and jam, composite optimizations, and architecture-specific directives (e.g., for generation of calls to SIMD intrinsics on the Blue Gene/P). On the right-hand side of Figure~\ref{fig:orio-example}, a separate tuning specification contains the information required for creating complete executable tests and running them, including variable types, initializations (e.g., random), the search heuristic (in this case, exhaustive search is indicated, other options are random, simplex, and simulated annealing), and execution environment details. When parallel resources are available, Orio executes multiple variants in the same parallel job (the granularity is indicated by the \texttt{num\_processes} field in the search portion of the tuning specification. At present the tuning specification must be created manually. When Orio is used in conjunction with compiler tools, such as the MATLAB compiler described in this paper, it should be possible to generate most of this tuning specification automatically.

@@

For the Xeon experiments, we used the Intel icc compiler (version 10.1) with the \texttt{-O3 -parallel} optimization flags. We generated and tested the following different implementations of each operation:
\begin{itemize}
\item C from MATLAB: C code generated by the MATLAB compiler described in Section~\ref{sec:matlab}.
\item BLAS: A version implemented using multiple BLAS calls and linked with the default BLAS library available on Ubutnu 8.04.
\item Intel MKL: The same code as the BLAS version, but linked to the Intel MKL libraries.
\item ATLAS: The same code as the BLAS version, but linked to the ATLAS libraries tuned for this machine.
\item Orio (Seq.): The C code generated by the MATLAB compiler annotated with sequential-only optimization directives.
\item Orio (Par.) : The C code generated by the MATLAB compiler annotated with both sequential and parallel optimization directives. This version was compiled with the \texttt{-O3 -openmp} options.
\end{itemize}

@@

The VADD operation is simply the addition of three arrays, storing the result in a fourth: $x = w + y + z$.
We optimized the performance for the VADD operation on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0. We measured the performance for two scenarios: using one core per processor and using all cores. In the single-core case, the codes were compiled with the following options: \texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot -qsmp=noauto}. For the multicore case, the only difference in the compilation was that for the non-Orio versions, the \texttt{-qsmp=auto} option was used.

@@

The extended Matlab code for the ATAX operation, $y = A' * (A * x)$, and the corresponding performance results are shown in Figure~\ref{fig:atax}. The Orio-tuned version that incorporates PLuTo-generated loop fusion optimizations and Orio unrolling and parallelization directives achieves the best performance for most problem sizes, outperforming the Intel MKL version by a factor of 2 to 5.7 and the compiler-optimized C version by a factor of 4 to 7.

@@

The extended MATLAB code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}. As in the ATAX example, the parallel Orio version achieves the best performance, although in this case the sequential Orio version performance almost as well.

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.3\textwidth}
\footnotesize
\begin{verbatim}
GEMVER
in
  A : column matrix,
  u1 : vector, u2 : vector,
  v1 : vector, v2 : vector,
  a : scalar, b : scalar,
  y : vector, z : vector
out
  B : column matrix,
  x : vector, w : vector
{
  B = A + u1 * v1' + u2 * v2'
  x = b * (B' * y) + z
  w = a * (B * x)
}
\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
\label{fig:gemver}
\end{figure}

@@

The extended Matlab code for the GESUMMV operation, $y = a * (A * x) + b * (B * x)$,
and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.

@@

The positive initial results from our three-pronged approach to generating tuned linear algebra routines motivate several future lines of investigation. At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions. We will next delve into each manual step and create the software that automates it and in effect serves as the ``glue'' between the different tools. This would enable application developers to easily produce custom linear algebra libraries tuned for specific architectures. We will also investigate extending the supported functionality of our MATLAB compiler beyond composed BLAS operations, thus benefiting a wider spectrum of applications.

Ultimately, the tuning process should result in more than one version of the computation since different versions are optimal for different problem sizes.

% BN: moved from search space section above
%Yotov:2005kx
We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations. The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.  Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.

\singlespacing
\bibliographystyle{siam}
\bibliography{paper}
\end{document}

>>>>

%, CGO 09: http://www.cgo.org/
% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
% Max 22 pages, double-spaced, 11pt or larger font
% Abstract 300 words: Sep. 12, 2008
% Full paper: Sep. 19, 2008

@@

\usepackage{url}
\usepackage{html}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{cite}
\usepackage{times}
\usepackage{amsmath}
\usepackage[margin=1in]{geometry}

\newenvironment{itemizer}{\begin{itemize}\setlength{\parsep}{0cm}\setlength{\itemsep}{-.3em}}{\end{itemize}}

@@

\maketitle
%
\onehalfspacing
\begin{abstract}
The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations. Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.
In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.  To generate highly optimized code from a high-level MATLAB prototype, we define a three-step approach.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.   We then use the polyhedral optimization system PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously. Finally, we annotate the resulting code with performance tuning directives and use the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment. Orio performs an automated empirical search to select the best among the multiple optimized code variants. 
We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
\end{abstract}
%
\doublespacing
%
\section{Introduction}
\label{sec:intro}

@@

Historically, the research community has pursued two separate paths toward the goal of making software run at near-peak levels.  The first of these builds on research into compilers and their associated technologies.  The holy grail of compilation research is to take an arbitrary code as input and produce optimal code as output for a given language and hardware platform.  Despite the immense effort that has been poured into that approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.  The second path is to identify kernel routines that constitute the dominant costs of a wide variety of applications.  When such kernels can be identified and agreed upon by the members of the community, programmers with the required level of technical knowledge can concentrate on producing these optimized kernel libraries for architectures of interest. A prime example is the high-performance BLAS libraries produced by a combination of hardware vendors, independent software vendors, and researchers.  Developers who write their codes calling these routines can achieve high performance across all supported architectures. The efforts in high-performance library development aptly demonstrate, however, that the research into compiler technologies has not been fully successful.  If compilers could take the relatively simple code even for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.    Like compiler optimization, the kernel library-oriented approach has significant limitations.  Support for a variety of architectures is the most serious problem because the optimizations necessary to achieve near-peak performance are by nature not portable.   Lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.  It is a sad fact of the computing industry that by the time highly-optimized code is available for a given architecture, that architecture is generally well on its way to obsolescence.

We are working on a combination of the two approaches designed to overcome some of their shortcomings.   This paper describes our multipronged approach toward the development of software infrastructure for \emph{generating} automatically tuned libraries for matrix algebra computations.  We briefly overview relevant prior and current research efforts in Section~\ref{sec:background}. In Section~\ref{sec:approach} we describe our MATLAB to C compiler and the empirical performance tuning system Orio and its use in conjunction with the PLuTo tool suite to generate and empirically evaluate many tuned versions of the C code generated by the MATLAB compiler.  In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.

@@

The most common approach  to tuning numerical codes is for an expert to transform the source manually, unrolling loops, blocking for multiple levels of cache, and inserting prefetch instructions.  The pitfalls of this approach are well-understood~\cite{Goedecker01}.  It requires a significant amount of time and effort. Optimizing code for one particular platform may in fact make it less efficient on other platforms and often makes it complex and hard to understand or maintain.    An alternative is the use of tuned libraries of key numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific computing in C++, or GPULib by Tech-X, an open source library for accelerating numerical computations on Graphics Processing Units~\cite{gpulib}, to name just a few). Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:approach}, a series of calls to library routines can incur high memory costs.

@@

The best performance today is obtained through approaches like that taken by PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists. Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated. Our efforts aim to capture this performance tuning expertise in a way that can be integrated into the application. We will thus enable performance portability and ensure that the improvements to the application are long-lasting.

A number of source or binary transformation tools for performance-improving optimizations exist. LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.  A relatively new tool, POET~\cite{POET} also supports a number of loop transformations. POET offers a complex template-based syntax for defining transformations in a language-independent manner. Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops. PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron. Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations. The transformations target both improved cache locality and parallelism.

\section{Our Approach to Optimizing Composed BLAS Operations}
\label{sec:approach}

%\section{Optimizing Composed Operations}
%\label{sec:composed}

Codes based on matrix algebra are generally constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency. Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.

The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS routines ~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.  Single routines that carry out more than one linear algebra operation are known as \emph{composed} BLAS. As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.   These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies. In the former case, the matrix $A$ is accessed twice, while in the latter it is accessed only once. Our preliminary results in prior research and in this paper indicate that this loop fusion leads to a routine that delivers significantly better performance than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for large matrix orders.
Composed routines are the focus of the work presented here, but much of what we discuss generalizes to a much broader array of computations.

@@

@@

\caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
\label{fig:process}
\end{figure}
To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin, we have developed a compiler that converts a MATLAB script into simple C code~\cite{Siek}.  After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} developed at Ohio State University to optimize for coarse-grained parallelism and locality simultaneously. Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters. Orio then performs an empirical search to select the best among the multiple optimized code variants.

%Probably need some words about WHY this multipronged approach is a good idea.

In the remainder of this section we describe each of the tools developed by the authors of this paper, namely the MATLAB to C compiler and the Orio empirical tuning tool.

@@

Orio~\cite{Norris:2007} is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions. Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.

@@

We use the VADD operation, $x = w + y + z$, as a simple example of the annotations and tuning specification used by Orio. Figure~\ref{fig:orio-example} shows the C code with annotation comments indicating that Orio should perform memory alignment and loop transformations. In addition to simple source transformation annotations in the example above, Orio supports a number of other transformations, including loop unroll and jam, composite optimizations, and architecture-specific directives (e.g., for generation of calls to SIMD intrinsics on the Blue Gene/P).

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.45\textwidth}
\scriptsize
\begin{verbatim}
void vadd(int n, double *y, double *x1,
               double *x2, double *x3)
{
    int i;

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.45\textwidth}
\scriptsize
\begin{verbatim}
spec align_unroll {
 def build  
 { 
   arg build_command = 'mpixlc -O3 -qstrict -lm'; 
 } 
 def performance_counter  
 { 
   arg method = 'basic timer'; 
   arg repetitions = 10000;
 } 
 def performance_params 
 { 
   param UF[] = range(1,20);
 } 
 def input_params 
 { 
   param SIZE = @THESIZE@;
   decl int n = SIZE;
   decl double x1[n] = random;   
   decl double x2[n] = random;   
   decl double x3[n] = random;   
   decl double y[n] = 0;
 } 
 def search
 {
   arg algorithm = 'Exhaustive';
   arg time_limit = 20;
   arg run_command = 'cqsub -n 64 -t 10 -q short ';
   arg num_processes = 64;
 }
}
\end{verbatim}
\end{minipage}\\
\end{tabular}
\caption{Orio example: Annotated C source code (left) and a sample tuning specification for the Blue Gene/P (right).}
\label{fig:orio-example}
\end{figure}

On the right-hand side of Figure~\ref{fig:orio-example}, we show a separate tuning specification example, which contains the information required for creating complete executable tests and running them, including variable types, initializations (e.g., random), the search heuristic (in this case, exhaustive search is indicated, other options are random, simplex, and simulated annealing), and execution environment details. When parallel resources are available, Orio executes multiple variants in the same parallel job (the granularity is indicated by the \texttt{num\_processes} field in the search portion of the tuning specification. At present the tuning specification must be created manually. When Orio is used in conjunction with compiler tools, such as the MATLAB compiler described in this paper, it should eventually be possible to generate most of this tuning specification automatically.

@@

For the Xeon experiments, we used the Intel icc compiler (version 10.1) with the \texttt{-O3 -parallel} optimization flags. We generated and tested the following different implementations of each operation:
\begin{itemizer}
\item C from MATLAB: C code generated by the MATLAB compiler described in Section~\ref{sec:matlab}.
\item BLAS: A version implemented using multiple BLAS calls and linked with the default BLAS library available on Ubutnu 8.04.
\item Intel MKL: The same code as the BLAS version, but linked to the Intel MKL libraries.
\item ATLAS: The same code as the BLAS version, but linked to the ATLAS libraries tuned for this machine.
\item Orio (Seq.): The C code generated by the MATLAB compiler annotated with sequential-only optimization directives.
\item Orio (Par.) : The C code generated by the MATLAB compiler annotated with both sequential and parallel optimization directives (note that the code is the same as for Orio (Seq.), the only difference is the value of a parallelization parameter in the separate tuning specification). This version was compiled with the \texttt{-O3 -openmp} options.
\end{itemizer}

@@

The VADD operation computes the sum of three arrays, storing the result in a fourth: $x = w + y + z$.
We optimized the performance for the VADD operation on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0. We measured the performance for two scenarios: using one core per processor and using all 4 cores on each processor. In the single-core case, the codes were compiled with the following options: \texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot -qsmp=noauto}. For the multicore case, the only difference in the compilation was that for the non-Orio versions, the \texttt{-qsmp=auto} option was used. The parallel Orio version differs from the sequential in that it contains parallelization directives, which result in the placement of OpenMP pragmas in the generated code.

@@

The extended Matlab code for the ATAX operation, $y = A' * (A * x)$, and the corresponding performance results are shown in Figure~\ref{fig:atax}. The Orio-tuned version that incorporates PLuTo-generated loop fusion optimizations and Orio parallelization directives achieves the best performance for most problem sizes, outperforming the Intel MKL version by a factor of 2 to 5.7 and the compiler-optimized C version by a factor of 4 to 7. The optimizations performed by both Orio versions include scalar replacement, vectorization, and unroll and jam.

@@

The extended MATLAB code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}. Here we used the same PLuTo and Orio optimizations as for the ATAX example and similarly the parallel Orio version achieves the best performance, although in this case the sequential Orio version performs almost as well suggesting that the compiler was not able to parallelize the code very effectively. For this operation, there are substantial performance differences between different BLAS versions, with the Intel MKL version achieving performance close to that of the simple compiler code.
\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.3\textwidth}
\footnotesize
\begin{verbatim}
GEMVER
in
  A : column matrix,
  u1 : vector, u2 : vector,
  v1 : vector, v2 : vector,
  a : scalar, b : scalar,
  y : vector, z : vector
out
  B : column matrix,
  x : vector, w : vector
{
  B = A + u1 * v1' + u2 * v2'
  x = b * (B' * y) + z
  w = a * (B * x)
}
\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
\label{fig:gemver}
\end{figure}

@@

The extended Matlab code for the GESUMMV operation, $y = a * (A * x) + b * (B * x)$,
and the corresponding performance results are shown in Figure~\ref{fig:gesummv}. Unlike all the other operations we tested, the best performing version is the ATLAS one, which substantially outperforms all the alternatives. The next best performance is exhibited by the parallel Orio version. This result motivates the development of transformations that enable the generation of code with calls to existing optimizing libraries in addition to the low-level loop transformations targeted by the current tools (PLuTo and Orio).

@@

We have described an approach to generating tuned linear algebra libraries from high-level annotated MATLAB code that involves a suite of tools to (1) translate the MATLAB code to C, (2) analyze the resulting loops and identify locatility and parallelism-enhancing optimizations using PLuTo, and (3) annotate the resulting C code with syntactic performance directives and use Orio to generate multiple optimized versions and empirically select the one with the best performance. Preliminary results from experiments with several composed BLAS operations show that in almost all cases, the optimized code generated by this suite of tools significantly outperforms the versions using tuned BLAS and aggressive compiler optimizations.

The positive initial results from our approach to generating tuned linear algebra routines motivate several future lines of investigation. At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions. We will work on interfacing our separately developed tools more closely and automating their interactions. This would enable application developers to easily produce custom linear algebra libraries tuned for specific architectures. We will also investigate extending the supported functionality of our MATLAB compiler beyond composed BLAS operations, thus benefiting a wider spectrum of applications.

Orio is growing in a number of directions. While it automates much of the tuning process, a number of steps still require nontrivial manual input. Orio's modular design allows new transformations to be added easily, which presents many opportunities for expanding the existing set of optimizations with new ones implemented from scratch or through interfaces to other tools, as we have successfully done with PLuTo. We are also working on generating code that contains special cases for different input parameter values since in many cases different optimizations are optimal for different input sizes.  We will also continue developing the parallel empirical search infrastructure; it is clear that more feedback between the MATLAB compiler and Orio will result in automation of the performance directive annotation process and a much more effective exploration of the search space. We plan to use a fast analytic cost function in the MATLAB compiler implementation to prune out regions of the search space that will not produce competitive implementations. The cost function is based on a model of the computer architecture, but does not necessarily account for all the details. Once the search space is narrowed, we will use the Orio empirical testing results with representative data sets to determine the exact performance of the alternatives and choose the best one.

Finally, while Orio currently generates various loop transformations, it does not generate calls to existing library implementations. For this paper, we implemented the BLAS versions of all operations manually. Our past success with identifying Blue Gene/L intrinsics from expression graphs motivate further research on identifying BLAS and other established library operations from the expression graphs of the MATLAB code.

\singlespacing
%% or stuff it in a footnote
\section*{Acknowledgments}
This work was supported by the Mathematical, Information, and
Computational Sciences Division subprogram of the Office of Advanced
Scientific Computing Research, Office of Science, U.S. Dept. of Energy,
under Contract DE-AC02-06CH11357. 
%TODO: need Colorado and Ohio acknowledgments

\bibliographystyle{acm}
\bibliography{paper}
